---
title: "EBOLA_ODEs"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This R Markdown document models the Seasonal Spillover of Ebola Virus using a system of coupled ordinary differential equations (ODEs).

```{r Load package, include=FALSE}
library(deSolve)
```
#########################################################
# Two Pacth EBOLA ODEs - No seasonality and no movement #
#########################################################
```{r ODEs, echo=FALSE}
###################################################################
# Disease Transmission Rate by (t) No seasonality and no movement #
###################################################################
#--------------------------------------------------------------------#
two_patch_SIR_no_seasonality <- function(time, variables, parameters) {
  with(as.list(c(variables, parameters)), {
    
    dS_A <- (mu_A * N_A) - ((beta_A)*(I_A/N_A)*(S_A)) - ((mu_A + phi_AB)*S_A) + (phi_BA*S_B) # Correct.
    
    dI_A <- (beta_A*(I_A/N_A)*S_A) - ((mu_A + gamma_A + phi_AB) * I_A) + (phi_BA*I_B) # Correct.
    
    dR_A <- (gamma_A*I_A)-((mu_A + phi_AB)*R_A) + (phi_BA*R_B)  # Correct.
    
#-------------------------------------------------------------------#
    
    dS_B <-(mu_B * N_B) - ((beta_B)*(I_B/N_B)*(S_B)) - ((mu_B + phi_BA)*S_B) + (phi_AB*S_A) # Correct.
    
    dI_B <-(beta_B*(I_B/N_B)*S_B) - ((mu_B + gamma_B + phi_BA)*I_B) + (phi_AB*I_A) # Correct.
    
    dR_B <-(gamma_B*I_B)-((mu_B + phi_BA)*R_B) + (phi_AB*R_A) # Correct.
    
    return(list(c(dS_A, dI_A, dR_A, dS_B, dI_B, dR_B)))
  })
}

# Set up initial values for the variables
initial_values <- c(
  S_A = 95, I_A = 5, R_A = 0,   # Initial values for patch A
  S_B = 995, I_B = 5, R_B = 0   # Initial values for patch B
)

# Define parameters for the system
parameters_values <- list(
  N_A = 100,
  N_B = 1000,
  mu_A = 0.06,
  mu_B = 0.02,
  gamma_A = 12,
  gamma_B = 12,
  beta_A = 60, # takes the average trans rate and amplitude of transmition
  beta_B = 5, # takes the average trans rate and amplitude of transmition
  phi_AB = 0, # takes the average disp rate and amplitude of dispertion
  phi_BA = 0 # takes the average disp rate and amplitude of dispertion
)

# Define a sequence of time points
time <- seq(0, 1, by = 0.01)

# Solve the system of differential equations
sir_values <- ode(
  y = initial_values,
  times = time,
  func = two_patch_SIR_no_seasonality,
  parms = parameters_values
)

# Print the solution
print(sir_values)
```
##################################
# No seasonality and no movement #
##################################
```{r ODEs, echo=FALSE}
#################################################
#            Transform to Data Frame            #
#################################################

sir_values <- as.data.frame(sir_values)
sir_values

######################################
#            PLOT PATCH A            #
######################################
with(sir_values, {
# plotting the time series of susceptibles:
  plot(time, S_A, type = "l", col = "blue",
       xlab = "Time", ylab = "Individuals", ylim = c(0, 100))
# adding the time series of infectious:
  lines(time, I_A, col = "red")
# adding the time series of recovered:
  lines(time, R_A, col = "green")
})

title("PATCH A")
# adding a legend:
legend("right", c("SA", "IA", "RA"),
       col = c("blue", "red", "green"), lty = 1, bty = "n")

#####################################
#            PLOT PATCH B           #
#####################################
with(sir_values, {
# plotting the time series of susceptibles:
  plot(time, S_B, type = "l", col = "blue",
       xlab = "Time", ylab = "Individuals", ylim = c(0, 1000))
# adding the time series of infectious:
  lines(time, I_B, col = "red")
# adding the time series of recovered:
  lines(time, R_B, col = "green")
})

title("PATCH B")
# adding a legend:
legend("right", c("SA", "IA", "RA"),
       col = c("blue", "red", "green"), lty = 1, bty = "n")

######################################
#            PLOT PATCH A I_A        #
######################################
with(sir_values, {
# plotting the time series of susceptibles:
  plot(time, I_A, col = "red", type = "l",
       xlab = "time", ylab = "IA", ylim = c(-0.5, 50))
})

title("PATCH A")
# adding a legend:

######################################
#            PLOT PATCH B I_B        #
######################################
with(sir_values, {
# plotting the time series of susceptibles:
  plot(time, I_B, col = "red", type = "line",
       xlab = "time", ylab = "IB", ylim = c(-0.5, 5))
})

# adding a legend:
title("PATCH B")
```
#########################################
# Two Pacth EBOLA ODEs with seasonality #
#########################################
```{r ODEs, echo=FALSE}
#####################################################
# Disease Transmission Rate by (t) with Seasonality #
#####################################################
# sigma          #  
seasonality=0.5  #
##################

beta_A<-function(time){
  return(60*(1+(seasonality*cos((2*pi*time)/1)))) # Correct.
}

beta_B<-function(time){
  
  return(5*(1+(seasonality*cos((2*pi*time)/1)))) # Correct.
         }
##########################
# Disease Dispersal Rate #  
##########################
# sigma          #  
seasonality=0.5  #
##################

phi_AB<-function(time){
  return(0.4*(1 + (seasonality*cos((2*pi*time/1))))) # Correct.
}

phi_BA<-function(time){
  
  return(0.04*(1 + (seasonality*(cos((2*pi*time)/1))))) # Correct.
         }
#--------------------------------------------------------------------#
two_patch_SIR <- function(time, variables, parameters, phi_BA, phi_AB, beta_A, beta_B) {
  with(as.list(c(variables, parameters)), {
    
    dS_A <- (mu_A * N_A) - ((beta_A(time))*(I_A/N_A)*(S_A)) - ((mu_A + phi_AB(time))*S_A) + (phi_BA(time)*S_B) # Correct.
    
    dI_A <- (beta_A(time)*(I_A/N_A)*S_A) - ((mu_A + gamma_A + phi_AB(time)) * I_A) + (phi_BA(time)*I_B) # Correct.
    
    dR_A <- (gamma_A*I_A)-((mu_A + phi_AB(time))*R_A) + (phi_BA(time)*R_B)  # Correct.
    
#-------------------------------------------------------------------#
    dS_B <-(mu_B * N_B) - ((beta_B(time))*(I_B/N_B)*(S_B)) - ((mu_B + phi_BA(time))*S_B) + (phi_AB(time)*S_A) # Correct.
    
    dI_B <-(beta_B(time)*(I_B/N_B)*S_B) - ((mu_B + gamma_B + phi_BA(time))*I_B) + (phi_AB(time)*I_A) # Correct.
    
    dR_B <-(gamma_B*I_B)-((mu_B + phi_BA(time))*R_B) + (phi_AB(time)*R_A) # Correct.
    
    return(list(c(dS_A, dI_A, dR_A, dS_B, dI_B, dR_B)))
  })
}

# Set up initial values for the variables
initial_values <- c(
  S_A = 95, I_A = 5, R_A = 0,   # Initial values for patch A
  S_B = 995, I_B = 5, R_B = 0   # Initial values for patch B
)

# Define parameters for the system
parameters_values <- list(
  N_A = 100,
  N_B = 1000,
  mu_A = 0.06,
  mu_B = 0.02,
  gamma_A = 12,
  gamma_B = 12,
  phi_BA = phi_BA,
  phi_AB = phi_AB,
  beta_A = beta_A,
  beta_B = beta_B
)

# Define a sequence of time points
time <- seq(0, 1, by = 0.01)

# Solve the system of differential equations
sir_values <- ode(
  y = initial_values,
  times = time,
  func = two_patch_SIR,
  parms = parameters_values
)

# Print the solution
print(sir_values)
```

```{r ODEs, echo=FALSE}
#################################################
#            Transform to Data Frame            #
#################################################

sir_values <- as.data.frame(sir_values)
sir_values

######################################
#            PLOT PATCH A            #
######################################
with(sir_values, {
# plotting the time series of susceptibles:
  plot(time, S_A, type = "l", col = "blue",
       xlab = "time", ylab = "number of people", ylim = c(0, 100))
# adding the time series of infectious:
  lines(time, I_A, col = "red")
# adding the time series of recovered:
  lines(time, R_A, col = "green")
})

# adding a legend:
legend("right", c("SA", "IA", "RA"),
       col = c("blue", "red", "green"), lty = 1, bty = "n")

#####################################
#            PLOT PATCH B           #
#####################################
with(sir_values, {
# plotting the time series of susceptibles:
  plot(time, S_B, type = "l", col = "blue",
       xlab = "time", ylab = "number of people", ylim = c(0, 1000))
# adding the time series of infectious:
  lines(time, I_B, col = "red")
# adding the time series of recovered:
  lines(time, R_B, col = "green")
})

# adding a legend:
legend("right", c("SB", "IB", "RB"),
       col = c("blue", "red", "green"), lty = 1, bty = "n")

######################################
#            PLOT PATCH A I_A        #
######################################
with(sir_values, {
# plotting the time series of susceptibles:
  plot(time, I_A, col = "red", type = "l",
       xlab = "time", ylab = "number of people", ylim = c(0, 65))
})

# adding a legend:
legend("right", c("infectious"),
       col = c("red"), lty = 1, bty = "n")

######################################
#            PLOT PATCH A I_B        #
######################################
with(sir_values, {
# plotting the time series of susceptibles:
  plot(time, I_B, col = "red", type = "l",
       xlab = "time", ylab = "number of people", ylim = c(0, 5.5))
})

# adding a legend:
legend("right", c("infectious"),
       col = c("red"), lty = 1, bty = "n")
```
########################################################
# Two Pacth EBOLA ODEs with seasonality  for 100 years #
########################################################
```{r ODEs, echo=FALSE}
#####################################################
# Disease Transmission Rate by (t) with Seasonality #
#####################################################
# sigma          #  
seasonality=0.5  #
##################

beta_A<-function(time){
  return(60*(1+(seasonality*cos((2*pi*time)/1)))) # Correct.
}

beta_B<-function(time){
  
  return(5*(1+(seasonality*cos((2*pi*time)/1)))) # Correct.
         }
##########################
# Disease Dispersal Rate #  
##########################
# sigma          #  
seasonality=0.5  #
##################

phi_AB<-function(time){
  return(0.4*(1 + (seasonality*cos((2*pi*time/1))))) # Correct.
}

phi_BA<-function(time){
  
  return(0.04*(1 + (seasonality*(cos((2*pi*time)/1))))) # Correct.
         }
#--------------------------------------------------------------------#
two_patch_SIR <- function(time, variables, parameters, phi_BA, phi_AB, beta_A, beta_B) {
  with(as.list(c(variables, parameters)), {
    
    dS_A <- (mu_A * N_A) - ((beta_A(time))*(I_A/N_A)*(S_A)) - ((mu_A + phi_AB(time))*S_A) + (phi_BA(time)*S_B) # Correct.
    
    dI_A <- (beta_A(time)*(I_A/N_A)*S_A) - ((mu_A + gamma_A + phi_AB(time)) * I_A) + (phi_BA(time)*I_B) # Correct.
    
    dR_A <- (gamma_A*I_A)-((mu_A + phi_AB(time))*R_A) + (phi_BA(time)*R_B)  # Correct.
    
#-------------------------------------------------------------------#
    dS_B <-(mu_B * N_B) - ((beta_B(time))*(I_B/N_B)*(S_B)) - ((mu_B + phi_BA(time))*S_B) + (phi_AB(time)*S_A) # Correct.
    
    dI_B <-(beta_B(time)*(I_B/N_B)*S_B) - ((mu_B + gamma_B + phi_BA(time))*I_B) + (phi_AB(time)*I_A) # Correct.
    
    dR_B <-(gamma_B*I_B)-((mu_B + phi_BA(time))*R_B) + (phi_AB(time)*R_A) # Correct.
    
    return(list(c(dS_A, dI_A, dR_A, dS_B, dI_B, dR_B)))
  })
}

# Set up initial values for the variables
initial_values <- c(
  S_A = 95, I_A = 5, R_A = 0,   # Initial values for patch A
  S_B = 995, I_B = 5, R_B = 0   # Initial values for patch B
)

# Define parameters for the system
parameters_values <- list(
  N_A = 100,
  N_B = 1000,
  mu_A = 0.06,
  mu_B = 0.02,
  gamma_A = 12,
  gamma_B = 12,
  phi_BA = phi_BA,
  phi_AB = phi_AB,
  beta_A = beta_A,
  beta_B = beta_B
)

# Define a sequence of time points
time <- seq(0, 100, by = 0.01)

# Solve the system of differential equations
sir_values <- ode(
  y = initial_values,
  times = time,
  func = two_patch_SIR,
  parms = parameters_values
)

# Print the solution
print(sir_values)
```
########################################################
# Two Pacth EBOLA ODEs with seasonality  for 100 years #
########################################################
```{r ODEs, echo=FALSE}
#################################################
#            Transform to Data Frame            #
#################################################

sir_values <- as.data.frame(sir_values)
sir_values

######################################
#            PLOT PATCH A            #
######################################
with(sir_values, {
# plotting the time series of susceptibles:
  plot(time, S_A, type = "l", col = "blue",
       xlab = "time", ylab = "number of people", ylim = c(0, 100))
# adding the time series of infectious:
  lines(time, I_A, col = "red")
# adding the time series of recovered:
  lines(time, R_A, col = "green")
})

# adding a legend:
legend("right", c("SA", "IA", "RA"),
       col = c("blue", "red", "green"), lty = 1, bty = "n")

#####################################
#            PLOT PATCH B           #
#####################################
with(sir_values, {
# plotting the time series of susceptibles:
  plot(time, S_B, type = "l", col = "blue",
       xlab = "time", ylab = "number of people", ylim = c(0, 1000))
# adding the time series of infectious:
  lines(time, I_B, col = "red")
# adding the time series of recovered:
  lines(time, R_B, col = "green")
})

# adding a legend:
legend("right", c("SB", "IB", "RB"),
       col = c("blue", "red", "green"), lty = 1, bty = "n")
```
########################################################
# Two Pacth EBOLA ODEs with seasonality  for 100 years #
########################################################
```{r ODEs, echo=FALSE}
######################################
#            PLOT PATCH A            #
######################################
with(sir_values, {
# plotting the time series of susceptibles:
  plot(time, S_A, type = "l", col = "blue",
       xlab = "time", ylab = "number of people", ylim = c(0, 100), xlim = c(75, 100))
# adding the time series of infectious:
  lines(time, I_A, col = "red")
# adding the time series of recovered:
  lines(time, R_A, col = "green")
})

#####################################
#            PLOT PATCH B           #
#####################################
with(sir_values, {
# plotting the time series of susceptibles:
  plot(time, S_B, type = "l", col = "blue",
       xlab = "time", ylab = "number of people", ylim = c(0, 1000), xlim=c(75,100))
# adding the time series of infectious:
  lines(time, I_B, col = "red")
# adding the time series of recovered:
  lines(time, R_B, col = "green")
})

# adding a legend:
legend("right", c("SB", "IB", "RB"),
       col = c("blue", "red", "green"), lty = 1, bty = "n")
```
```{r ODEs, echo=FALSE}
######################################
#            PLOT PATCH A I_A        #
######################################
with(sir_values, {
# plotting the time series of susceptibles:
  plot(time, I_A, col = "red", type = "l",
       xlab = "time", ylab = "number of people", ylim = c(0, 5), xlim=c(75,100))
})

# adding a legend:
legend("right", c("infectious"),
       col = c("red"), lty = 1, bty = "n")

######################################
#            PLOT PATCH A I_B        #
######################################
with(sir_values, {
# plotting the time series of susceptibles:
  plot(time, I_B, col = "red", type = "l",
       xlab = "time", ylab = "number of people", ylim = c(0, 0.3), xlim=c(75,100))
})

# adding a legend:
legend("right", c("infectious"),
       col = c("red"), lty = 1, bty = "n")
```